---
title: "Assignment 3"
author: "Branden Lee"
date: "16/04/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 1

```{r}
set.seed(0)

# Number of simulations
N <- 10000

# Number of bootstrap samples per simulation
B <- 1000

# Function for calculating the statistic of interest
T <- function(sample) (quantile(sample, .75, names=FALSE) - 
                         quantile(sample, .25, names=FALSE)) / 1.34

# True value of the statistic
true_value <- (qt(.75, df=3) - qt(.25, df=3)) / 1.34

# Returns true if x is in the interval [l, r]
covered <- function(x, l, r) x >= l & x <= r

# Calculates a normal interval
normal_ci <- function(mean, sd) mean + c(-1, 1) * qnorm(0.975) * sd

# Calculates proportion of intervals that contain the true value of the statistic
coverage_probability <- function(intervals) {
  mean(apply(intervals, MARGIN=1, FUN=function(x) covered(true_value, x[1], x[2])))
}

# Calculates average length of intervals
average_length <- function(intervals) {
  mean(apply(intervals, MARGIN=1, FUN=function(x) x[2] - x[1]))
}


normal_jackknife <- function(num_sims) {
  est <- c()
  se <- c()
  for (N in 1:num_sims) {
    s <- rt(25, 3)
    
    loo_est <- c()
    for (i in 1:length(s)) {
      loo_est <- c(loo_est, T(s[-i]))
    }
    est <- c(est, T(s))
    se <- c(se, sqrt(24 * mean((loo_est - mean(loo_est)) ^ 2)))
  }
  return(cbind(est, se))
}
output_a <- normal_jackknife(N)

est_a <- output_a[,1]
se_a <- output_a[,2]

ci_a <- t(apply(cbind(est_a, se_a), MARGIN=1, FUN=function(x) normal_ci(x[1], x[2])))

coverage_probability_a <- coverage_probability(ci_a)
average_length_a <- average_length(ci_a)


normal_bootstrap <- function(num_sims, B) {
  est <- c()
  se <- c()
  for (N in 1:num_sims) {
    s <- rt(25, 3)

    bootstrap_est <- c()
    for (b in 1:B) {
      bootstrap_sample <- sample(s, size=25, replace=TRUE)
      bootstrap_est <- c(bootstrap_est, T(bootstrap_sample))
    }
    theta_hat <- T(s)
    est <- c(est, theta_hat)
    se <- c(se, sqrt(sum((bootstrap_est - theta_hat) ^ 2) / (B - 1)))
  }
  return(cbind(est, se))
}
output_b <- normal_bootstrap(N, B)

est_b <- output_b[,1]
se_b <- output_b[,2]

ci_b <- t(apply(cbind(est_b, se_b), MARGIN=1, FUN=function(x) normal_ci(x[1], x[2])))

coverage_probability_b <- coverage_probability(ci_b)
average_length_b <- average_length(ci_b)


percentile_bootstrap <- function(num_sims, B) {
  intervals <- c()
  for (N in 1:num_sims) {
    s <- rt(25, 3)
    bootstrap_est <- c()
    for (b in 1:B) {
      bootstrap_sample <- sample(s, size=25, replace=TRUE)
      bootstrap_est <- c(bootstrap_est, T(bootstrap_sample))
    }
    intervals <- rbind(intervals, quantile(bootstrap_est, c(.025, .975), names=FALSE))
  }
  intervals
}
ci_c <- percentile_bootstrap(N, B)

coverage_probability_c <- coverage_probability(ci_c)
average_length_c <- average_length(ci_c)
```


```{r echo=FALSE}
format_interval <- function(x) {
  paste0('(', round(x[1], 5), ', ', round(x[2], 5), ')')
}

knitr::kable(
rbind(c('a) Jackknife Normal', coverage_probability_a, round(average_length_a, 5)),
      c('b) Bootstrap Normal', coverage_probability_b, round(average_length_b, 5)),
      c('c) Bootstrap Percentile', coverage_probability_c, round(average_length_c, 5))),
col.names=c('Interval', 'Coverage Probability', 'Average Length'),
caption="Resampling Simulations",
digits=5
  )
```
The bootstrap normal interval results in a coverage probability closest to the the desired confidence level, while both the jackknife normal and bootstrap percentile deviate from this confidence level. The significantly smaller coverage probability and average length using the normal jackknife interval in comparison to the other methods suggests that this method is underestimating the variance.