---
title: "Assignment 3"
author: "Branden Lee"
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document: default
header-includes: \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# library(survey)
library(Matching)
```

### Question 1

```{r}
# Setup
set.seed(1)

# Some useful functions
expit <- function(x) {exp(x) / (1 + exp(x))}

eta <- function(x1,x2,a) {exp(0.5 + 4*expit(3*x1) + x2 + 0.5*a)}

bias <- function(x) {sum(x-0.5) / length(x)}

stddev <- function(x) {sqrt(sum((x-mean(x))^2)/(length(x)-1))}

mse <- function(x) {sum((x-0.5)^2) / length(x)}


# Number of simulations
k=1000


# Creating empty vectors to store estimates
matchingEstimate1 <- rep(NA, k)
matchingEstimate2 <- rep(NA, k)
matchingEstimate3 <- rep(NA, k)
matchingEstimate4 <- rep(NA, k)

ipwEstimate1 <- rep(NA, k)
ipwEstimate2 <- rep(NA, k)
ipwEstimate3 <- rep(NA, k)
ipwEstimate4 <- rep(NA, k)

stratifiedEstimate1 <- rep(NA, k)
stratifiedEstimate2 <- rep(NA, k)
stratifiedEstimate3 <- rep(NA, k)
stratifiedEstimate4 <- rep(NA, k)


# Loop for simulation
for(i in 1:k) {
  # Create dataframe and generate covariates
  df <- data.frame(matrix(rnorm(7500), ncol=3))
  colnames(df) <- c('x1', 'x2', 'x3')

  
  # True propensity score
  df$ps <- expit(df$x1 * 0.5 + df$x3 * 0.75)

  
  # Treatment assignment
  df$A <- apply(df[,c('x1','x3')],
                MARGIN=1,
                FUN=function(x) rbinom(1, 1, expit(x['x1'] * 0.5 + x['x3'] * 0.75)))
  
  
  # Count outcome
  df$Y <- apply(df[,c('x1','x2','A')],
                MARGIN=1,
                FUN=function(x) rpois(1, lambda=eta(x[1],x[2],x[3])))

  
  # Fit propensity score models
  df$ps1 <- predict(glm(A~x1, family=binomial, data=df), type='response')
  df$ps2 <- predict(glm(A~x1+x3, family=binomial, data=df), type='response')
  df$ps3 <- predict(glm(A~x1+x2, family=binomial, data=df), type='response')
  df$ps4 <- predict(glm(A~x1+x2+x3, family=binomial, data=df), type='response')
  
  
  # Matching estimates
  
  # Generic function to calculate estimated causal log rate ratio by matching the
  # provided estimated propensity scores using 1-1 matching with replacement and no
  # ties.
  estimateMatching <- function(.ps) {
    matching <- Match(Y=df$Y, Tr=df$A, X=.ps, M=1, replace=TRUE, ties=FALSE, 
                      estimand='ATE', caliper=.2)
    matched <- df[c(matching$index.treated, matching$index.control),]
    glm(Y~A, family=poisson, data=matched)$coef[2]
  }
  matchingEstimate1[i] <- estimateMatching(df$ps1)
  matchingEstimate2[i] <- estimateMatching(df$ps2)
  matchingEstimate3[i] <- estimateMatching(df$ps3)
  matchingEstimate4[i] <- estimateMatching(df$ps4)

  # matching1 <- Match(Y=df$Y, Tr=df$A, X=df$ps1, replace=TRUE, ties=FALSE, estimand='ATE',
  #                    caliper=.2)
  # matched1 <- df[c(matching1$index.treated, matching1$index.control),]
  # matchingEstimate1[i] <- glm(Y~A, family=poisson, data=matched1)$coef[2]
  # print(matchingEstimate1[i])
  # 
  # matching2 <- Match(Y=df$Y, Tr=df$A, X=df$ps2, M=1, replace=TRUE, ties=FALSE, estimand='ATE',
  #                    caliper=.2)
  # matched2 <- df[c(matching2$index.treated, matching2$index.control),]
  # matchingEstimate2[i] <- glm(Y~A, family=poisson, data=matched2)$coef[2]
  # 
  # matching3 <- Match(Y=df$Y, Tr=df$A, X=df$ps3, M=1, replace=TRUE, ties=FALSE, estimand='ATE',
  #                    caliper=.2)
  # matched3 <- df[c(matching3$index.treated, matching3$index.control),]
  # matchingEstimate3[i] <- glm(Y~A, family=poisson, data=matched3)$coef[2]
  # 
  # matching4 <- Match(Y=df$Y, Tr=df$A, X=df$ps4, M=1, replace=TRUE, ties=FALSE, estimand='ATE',
  #                    caliper=.2)
  # matched4 <- df[c(matching4$index.treated, matching4$index.control),]
  # matchingEstimate4[i] <- glm(Y~A, family=poisson, data=matched4)$coef[2]

  
  # IPW estimates
  
  # Returns adjusted weights (set to 20 if larger then 20) based on given propensity 
  # score model.
  adjustedWeights <- function(.ps) {
    w <- rep(NA, nrow(df))
    w[df$A==1] <- 1 / .ps[df$A==1]
    w[df$A==0] <- 1 / (1 - .ps[df$A==0])
    ifelse(w>20, 20, w)
  }
  ipwEstimate1[i] <- glm(Y~A, family=poisson, weights=adjustedWeights(df$ps1), 
                         data=df)$coef[2]
  ipwEstimate2[i] <- glm(Y~A, family=poisson, weights=adjustedWeights(df$ps2), 
                         data=df)$coef[2]
  ipwEstimate3[i] <- glm(Y~A, family=poisson, weights=adjustedWeights(df$ps3), 
                         data=df)$coef[2]
  ipwEstimate4[i] <- glm(Y~A, family=poisson, weights=adjustedWeights(df$ps4), 
                         data=df)$coef[2]

  
  # Stratification
  
  # Generic function for estimating causal log rate ratio by stratifying on provided
  # estimated propensity scores.
  estimateStratified <- function(x) {
    df$strata <- cut(x, breaks=quantile(df$ps1, probs=(seq(0, 1, 0.2))),
                 labels=1:5, include.lowest=TRUE)
    mean(sapply(1:5, 
                FUN=function(z) glm(Y~A, family=poisson, data=df[df$strata==z,])$coef[2]))
  }
  
  stratifiedEstimate1[i] <- estimateStratified(df$ps1)
  stratifiedEstimate2[i] <- estimateStratified(df$ps2)
  stratifiedEstimate3[i] <- estimateStratified(df$ps3)
  stratifiedEstimate4[i] <- estimateStratified(df$ps4)
    
  
  # df$strata <- cut(df$ps1, breaks=quantile(df$ps1, probs=(seq(0, 1, 0.2))),
  #                labels=c(1,2,3,4,5), include.lowest=TRUE)
  # stratifiedEstimate1[i] <- mean(sapply(1:5, FUN=function(x)
  #   glm(Y~A, family=poisson, data=df[df$strata==x,])$coef[2]))
  # df$strata <- cut(df$ps2, breaks=quantile(df$ps2, probs=(seq(0, 1, 0.2))),
  #                labels=c(1,2,3,4,5), include.lowest=TRUE)
  # stratifiedEstimate2[i] <- mean(sapply(1:5, FUN=function(x)
  #   glm(Y~A, family=poisson, data=df[df$strata==x,])$coef[2]))
  # df$strata <- cut(df$ps3, breaks=quantile(df$ps3, probs=(seq(0, 1, 0.2))),
  #                labels=c(1,2,3,4,5), include.lowest=TRUE)
  # stratifiedEstimate3[i] <- mean(sapply(1:5, FUN=function(x)
  #   glm(Y~A, family=poisson, data=df[df$strata==x,])$coef[2]))
  # df$strata <- cut(df$ps4, breaks=quantile(df$ps4, probs=(seq(0, 1, 0.2))),
  #                labels=c(1,2,3,4,5), include.lowest=TRUE)
  # stratifiedEstimate4[i] <- mean(sapply(1:5, FUN=function(x)
  #   glm(Y~A, family=poisson, data=df[df$strata==x,])$coef[2]))
}

```

\begin{center}
\captionof{table}{1-1 Matching with Replacement}
\begin{tabular}{|l|c|c|c|}
\hline
& Bias & Standard Deviation & Mean Squared Error\\
\hline
$PS_1$ & `r bias(matchingEstimate1)` & `r stddev(matchingEstimate1)` & `r mse(matchingEstimate1)`\\
$PS_2$ & `r bias(matchingEstimate2)` & `r stddev(matchingEstimate2)` & `r mse(matchingEstimate2)`\\
$PS_3$ & `r bias(matchingEstimate3)` & `r stddev(matchingEstimate3)` & `r mse(matchingEstimate3)`\\
$PS_4$ & `r bias(matchingEstimate4)` & `r stddev(matchingEstimate4)` & `r mse(matchingEstimate4)`\\
\hline
\end{tabular}
\end{center}

\begin{center}
\captionof{table}{IPW}
\begin{tabular}{|l|c|c|c|}
\hline
& Bias & Standard Deviation & Mean Squared Error\\
\hline
$PS_1$ & `r bias(ipwEstimate1)` & `r stddev(ipwEstimate1)` & `r mse(ipwEstimate1)`\\
$PS_2$ & `r bias(ipwEstimate2)` & `r stddev(ipwEstimate2)` & `r mse(ipwEstimate2)`\\
$PS_3$ & `r bias(ipwEstimate3)` & `r stddev(ipwEstimate3)` & `r mse(ipwEstimate3)`\\
$PS_4$ & `r bias(ipwEstimate4)` & `r stddev(ipwEstimate4)` & `r mse(ipwEstimate4)`\\
\hline
\end{tabular}
\end{center}

\begin{center}
\captionof{table}{Stratification into quintiles}
\begin{tabular}{|l|c|c|c|}
\hline
& Bias & Standard Deviation & Mean Squared Error\\
\hline
$PS_1$ & `r bias(stratifiedEstimate1)` & `r stddev(stratifiedEstimate1)` & `r mse(stratifiedEstimate1)`\\
$PS_2$ & `r bias(stratifiedEstimate2)` & `r stddev(stratifiedEstimate2)` & `r mse(stratifiedEstimate2)`\\
$PS_3$ & `r bias(stratifiedEstimate3)` & `r stddev(stratifiedEstimate3)` & `r mse(stratifiedEstimate3)`\\
$PS_4$ & `r bias(stratifiedEstimate4)` & `r stddev(stratifiedEstimate4)` & `r mse(stratifiedEstimate4)`\\
\hline
\end{tabular}
\end{center}


Based on the simulation, propensity score model 2 ($x_1, x_2$) results in the best estimated causal log rate ratio in terms of both standard deviation and mean squared error, while model 3 ($x_1, x_3$) performs the worst despite being the true propensity score model. This agrees with the fact that it is optimal to choose a propensity score model that includes only the covariates marginally related with the outcome.