---
title: "STAT831 Assignment 2"
author: "Branden Lee 20877653"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 2
##### a)
```{r}
library(GLMsData)
data(germ)

germ$resp <- cbind(germ$Germ, germ$Total - germ$Germ)
binomialModel1 <- glm(resp ~ Seeds * Extract, family = binomial, data = germ)

beta <- exp(binomialModel1$coefficients)
```
The odds of OA73 bean seeds germinating is `r beta[1]`.  
The odds of OA75 bean seeds germinating versus the odds of OA73 bean seeds germinating is  `r beta[2]`.  
The odds of OA73 cucumber seeds germinating versus the odds of OA73 bean seeds is `r beta[3]`.  
The estimate $e^{\hat{\beta}_3} = `r beta[4]`$ suggests that the germination of cucumber seeds may depend on the type of seed.


##### b)
```{r}
data(germBin)

germBin$ResultF <- factor(germBin$Result, levels = c('NotGerm', 'Germ'))
binaryModel1 <- glm(ResultF ~ Seeds * Extract, family=binomial, data=germBin)

summary(binomialModel1)
summary(binaryModel1)
```
It's clear from the summaries that the estimates for the coefficients agree and the deviances differ.

##### c)
```{r}
binomialModel2 <- glm(resp ~ Seeds + Extract, family = binomial, data = germ)
binomialDeviance <- binomialModel2$deviance - binomialModel1$deviance
binomialDevianceDf <- binomialModel2$df.residual - binomialModel1$df.residual

binaryModel2 <- glm(ResultF ~ Seeds + Extract, family=binomial, data=germBin)
binaryDeviance <- binaryModel2$deviance - binaryModel1$deviance
binaryDevianceDf <- binaryModel2$df.residual - binaryModel1$df.residual
```
The hypothesis test of interest is  
$$H_0: \beta_3 = 0,\ H_a: \beta_3 \neq 0$$ 
Binomial model: Under $H_0$, $\Delta D\  \dot{\sim} \ \chi^2_{`r binomialDevianceDf`}$:
$$\Delta D = D_0 - D_a = `r binomialModel2$deviance` - `r binomialModel1$deviance` = `r binomialDeviance`$$  
Binary model: Under $H_0$, $\Delta D\ \dot{\sim} \ \chi^2_{`r binaryDevianceDf`}$:
$$\Delta D = D_0 - D_a = `r binaryModel2$deviance` - `r binaryModel1$deviance` = `r binaryDeviance`$$
The above agrees with the derivation in question 1c i.e. that the deviance test statistics for testing nested models are equivalent.
