---
title: "STAT831: Assignment 1"
author: "Branden Lee 20877653"
header-includes: \usepackage{amsmath}
geometry: margin=2cm
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 3
(a) $f(y; \mu , \lambda) = \sqrt{\frac{\lambda}{2 \pi y^3}} exp{\frac{-\lambda(y - \mu)^2}{2 y \mu^2}}
= exp \left( \frac{\frac{1}{\mu^2}y-\frac{2}{\mu}}{-\frac{2}{\lambda}} - \frac{\lambda}{2y} + \frac{1}{2} log \frac{\lambda}{2 \pi y^3} \right)$

Taking $\theta = \frac{1}{\mu^2}, \;\: \phi = \frac{1}{\lambda}, \;\: a(\phi) = -2\phi, \;\: b(\theta) = 2\sqrt{\theta}, \;\: c(y; \phi) = - \frac{\lambda}{2y} + \frac{1}{2} log \frac{\lambda}{2 \pi y^3}$, we see that $Y$ is a member of the exponential family.

(b) $E(Y) = b'(\theta) = \theta^{-\frac{1}{2}} = \mu$  
$Var(Y) = a(\phi)b''(\theta) = -2 \phi (-\frac{1}{2}\theta^{-\frac{3}{2}}) = \frac{\mu^3}{\lambda}$  
$V(\mu) = b''(\theta) = -\frac{1}{2} \theta^{-\frac{3}{2}} = -\frac{1}{2}\mu^3$


(c) $\theta = \frac{1}{\mu^2} = \eta \implies g(\mu) = \frac{1}{\mu^2}$ is the canonical link function.

(d)
```{r, tidy=TRUE}
dinvgauss <- function(y, mu, lambda)
  {(lambda / (2 * pi * y^3)) ^ 0.5 * exp(-lambda * (y - mu) ^ 2 / (2 * y * mu ^2))}
density1 <- function(y){dinvgauss(y, 10, 5)}
curve(density1, 0, 40, xlim=c(0,40), ylim=c(0,0.17),type='l', lty=1, las = 1,
      main=expression(paste('Density of Inverse Gaussian for various ', mu, ', ', lambda)), 
      col = 'grey', xlab='y', ylab=expression(paste('f(y; ', mu, ', ',lambda, ')')))
density2 <- function(y){dinvgauss(y, 10, 10)}
curve(density2, 0, 40, type='l', col='blue', lty=2, add=TRUE)
density3 <- function(y){dinvgauss(y, 10, 20)}
curve(density3, 0, 40, type='l', col='black', lty=3, add=TRUE)
density4 <- function(y){dinvgauss(y, 20, 10)}
curve(density4, 0, 40, type='l', col='red', lty=4, add=TRUE)
legend("topright", legend=c(expression(paste(mu, '=10   ', lambda, '=5')),
                            expression(paste(mu, '=10   ', lambda,'=10')), 
                            expression(paste(mu, '=10   ', lambda, '=20')), 
                            expression(paste(mu, '=20   ', lambda, '=10'))), 
       col=c('grey', 'blue', 'black', 'red'), lty=c(1,2,3,4), bty='n')
```

The inverse Gaussian distribution would be appropriate for data that is continuous and
strictly non-negative.
