---
title: "STAT831: Assignment 1"
author: "Branden Lee 20877653"
header-includes: \usepackage{amsmath}
geometry: margin=2cm
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 1

```{r, tidy=TRUE}
beta <- c(100.812, 0.332, 0.411, -3.003, -0.362, -0.521)
se_beta <- c(13.096, 0.062, 0.090, 1.758, 2.732, 0.262)
p_val <- format(pt(beta / se_beta, 568, lower.tail=FALSE), scientific=TRUE) 
row_names <- c("Constant", "Age (in years)", "Waist circumference (cm)", 
               paste0("Alcohol ", "(yes: 1;no: 0)"), "Smoking (yes: 1;no: 0)", "Ambient temperature $(^{\\circ} C)$")
col_names <- c("$\\hat{\\beta_i}$", "$se(\\hat{\\beta_i})$", "$p_{val}$")
data <- matrix(c(beta, se_beta, p_val), ncol = 3)
df <- data.frame(data, row.names=row_names)
colnames(df) <- col_names
```

(a) Under the null hypothesis $H_0: \beta_0 = 0$, the p value is
\begin{align*}
p & = P(t_{574-5-1} > \tfrac{\hat{\beta_i}}{se(\hat{\beta_i)}}) \\
& = P(t_{568} > \tfrac{`r beta[1]`}{`r se_beta[1]`}) \\
& = `r p_val[1]`
\end{align*}
So we reject the null hypothesis. Similar to $\beta_0$, we reject the null hypotheses at the 95% confidence level corresponding to $\beta_1$ and $\beta_2$ for these respectively i.e. there is strong evidence that there is a relationship between age and systolic blood pressure as well as a relationship between waist circumference and systolic blood pressure (table below). However, since the p values for $\beta_3, \beta_4, \beta_5$ are larger than .05, we do not reject their respective null hypotheses at the 95% confidence level.
```{r echo=FALSE}
knitr::kable(df, escape=FALSE, scientific=TRUE)
```

(b) After adjusting for age, waist circumference, alcohol consumption and smoking habits, systolic blood pressure decreases by .521mm Hg for every degree Celsius increase in the temperature. 

(c)
```{r, tidy=TRUE}
x_0 <- c(1, 30, 100, 1, 0)
intercept_c <- sum(x_0 * beta[1:5])
eq_c = function(x){intercept_c + beta[6] * x}
curve(eq_c, 15, 50, type = 'l', lty = 2, xlab = bquote('Ambient temperature (' *degree ~ 'C)'),
       ylab = 'Systolic blood pressure (mm Hg)', 
      main=paste0("Ambient " , "temperature vs Systolic Blood Pressure"), las=1)
```

The relationship between ambient temperature and systolic blood pressure for 30 year old men who do not smoke, drink alcohol and have a waist circumference of 100 cm is modeled by the equation:
$$ Y = `r intercept_c``r beta[6]`x_5 $$
(d)
```{r}
lower <- beta[6] - qt(.975, 568) * se_beta[6]
upper <- beta[6] + qt(.975, 568) * se_beta[6]
```
A 95% confidence interval for the regression parameter for ambient temperature is given by the range of values between     $\hat{\beta_6} \pm  t_{568}(.975) se(\hat{\beta_6})$. The confidence interval is thus $(`r lower`, `r upper`)$.

(e)
```{r}
intercept_e <- sum(c(1, 45, 100, 1, 0) * beta[1:5])
eq_e <- function(x){intercept_e + beta[6] * x}
print(intercept_e)
```
The predicted mean systolic blood pressure for 45 year Ghanian men, don't smoke, drink alcohol and have a waist circumference
of 100 cm when the ambient temperature is $30^{\circ} C$ is
\begin{align*}
\widehat{E(Y|X = x_o)} & = `r intercept_e` - `r abs(beta[6])`x \\
& = `r intercept_e` - `r abs(beta[6])`\cdot 30 \\
& = `r eq_e(30)`
\end{align*}