---
title: "STAT831: Assignment 1 Supplement"
author: "Branden Lee 20877653"
header-includes: \usepackage{amsmath}
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 5

a.
The second order Taylor approximation of $\ell(\theta) = \ell(\theta, y)$ about the MLE $\hat{\theta}$ is given by
\begin{align*}
\ell(\theta) & \approx \ell(\hat{\theta}) + l'(\hat{\theta})(\theta - \hat{\theta}) + \frac{1}{2} l''(\hat{\theta})(\theta - \hat{\theta})^2 \\
& = \ell(\hat{\theta}) + S(\hat{\theta})(\theta - \hat{\theta}) - \frac{1}{2} I(\hat{\theta})(\theta - \hat{\theta})^2 \\
& = \ell(\hat{\theta}) - \frac{1}{2} I(\hat{\theta})(\theta - \hat{\theta})^2   && \text{since $S(\hat\theta) = 0$}
\end{align*}

b. $\ell(\theta) \approx \ell(\hat{\theta}) - \frac{1}{2} I(\hat{\theta})(\theta - \hat{\theta})^2 \iff \ell(\theta) - \ell(\hat{\theta}) \approx - \frac{1}{2} I(\hat{\theta})(\theta - \hat{\theta})^2 \iff (\hat{\theta} - \theta)^2 I(\hat{\theta}) \approx -2 (\ell(\theta) - \ell(\hat{\theta}))$

Since a Taylor approximation is only valid for $\left| \theta - \hat{\theta} \right|$ small, it follows that the Wald statistic is a good approximation to a likilhood ratio test when $\left| \theta - \hat{\theta} \right|$ is small.

c. As seen below, the plots reaffirm the derivation in part a since for values of $\hat{\lambda}$ near the MLE, the graph of the Wald statistic is an excellent approximation of the LR statistic. As $n$ gets larger, the interval over which the approximation is good becomes smaller.
```{r, tidy=TRUE}
n <- c(10, 25, 100)

lr <- function(lmbd){2 * (num / lmbd + num * log(lmbd) - num)}
wald <- function(lmbd){num * (1 - 1 / lmbd)^2}
for (num in n) {
title <- bquote("Likelihood based statistics for Exp(" ~ lambda ~ "), " ~ "n ="~ .(num))
curve(lr, 0.4, 2.1, main = title, type = 'l', lty = 1, col = 'grey', xlab = expression(widehat(lambda)), ylab = expression(paste("T(", widehat(lambda), ")")), xlim=c(0.5,2), ylim=c(0,50))
curve(wald, 0.4, 2.1, type = 'l', lty = 2, col = 'blue', add=TRUE)
legend('topright', legend = c("LR", 'Wald'), col = c('grey', 'blue'), lty=c(1,2), bty='n')
difference <- function(f_1, f_2, x) {abs(f_1(x) - f2(x))}

}
```
