---
title: "STAT840 A1 Q2"
author: "Branden Lee 20877653 Graduate Student"
date: "14/01/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### a)
$L(\theta|x=2.45) = \frac{1}{\sqrt{2 \pi}} e^{-\frac{(2.45-\theta)^2}{2}}$

#### b)

$L(\theta|x\in(0.9,4)) = P(0.9<x<4; \theta) = \int_{0.9}^4 \frac{1}{\sqrt{2 \pi}} e^{\frac{(x-\theta)^2}{2}} dx$

#### c)
Knowing the range of values that $x$ is in but not the exact value leads to a likelihood that seems to be uniformly larger than knowing $x=2.45$: this makes sense since knowing $x=2.45$ greatly limits the values of $\theta$ that could realistically result in sampling that particular value of $x$, while the uncertainty of $x\in(0.9,4)$ means a larger range of $\theta$ values could have resulted in such a sample.
```{r}
likelihood_1 <- function(theta) exp(-.5 * (2.45-theta)^2)/sqrt(2*pi)

f <- function(x, theta) exp(- .5 * (x - theta) ^ 2) / sqrt(2 * pi)

likelihood_2 <- function(theta) integrate(f, lower=0.9, upper=4, theta)$value
likelihood_2v <- Vectorize(likelihood_2)

curve(likelihood_1, from=-2.5, to=7.5, xlab=expression(theta),
      ylab=expression(paste('L(', theta, '|x)')), ylim=c(0,1), col='red', lty=2)
curve(likelihood_2v, from=-2.5, to=7.5, add=TRUE, col='blue', lty=2)
legend(-2.2, .95, legend=c('x=2.45', '0.9<x<4'), 
       col=c('red', 'blue'), lty=2:2)
```

