---
title: "STAT840 Final Exam"
author: "Branden Lee"
date: "4/26/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 3

#### a)
```{r}
set.seed(0)

p <- function(x) 1 / (pi * (1 + x ^ 2))

metropolis <- function(init, n) {
  x <- init
  
  sample <-  c(x)
  while (length(sample) < n) {
    proposal <- rnorm(1, mean=x, sd=1)
    r <- p(proposal) / p(x)
    if (r > 1 || runif(1) <= r) {x <- proposal}
    sample <- c(sample, x)
  }
  sample
}

output1 <- metropolis(3, 1000)
```
The steps for generating a sample from $\pi(x)$ using a Metropolis algorithm with $N(x,1)$ as the proposal density is as follows:  
1. Start with initial point $x_0$  
2. Sample proposal from $x\sim N(x_k, 1)$  
3. Compute $r=\frac{\pi(x)}{\pi(x_k)}$  
4. If $r>1$ set $x_{k+1}=x$. Else sample $u\sim Unif(0,1)$: if $u\leq r$ then set $x_{k+1}=x$  
5. Otherwise set $x_{k+1}=x_k$  
6. Set $k=k+1$
6. Go back to step 2 and repeat until desired sample size is obtained  

If the initial point is chosen to be close to the median $x=0$, the Metropolis algorithm seems to stabilise almost instantaneoulsy and certainly well within 1000 iterations. If it is chosen to be further away it still converges relatively quickly.
The Cauchy density was plotted on top of the relative frequency histogram in which the first 200 observations were discarded. We can see that the sample is quite good.

```{r, echo=FALSE}
plot(x=1:length(output1), y=output1, type='l', xlab='t',
     ylab=expression(X[t]), main='Time Series Plot')
hist(output1[201:1000],  xlab='x', prob=TRUE, main='Relative Frequency')
curve(dcauchy, add=TRUE)
```

#### b)
```{r}
sample <- output1[201:1000]
med <- median(sample)
se <- pi * (1 + med ^ 2) / (2 * sqrt(length(sample)))
med # Median
se # Standard error calculated using the formula provided
```
For good measure, the first 200 samples were discarded and the remaining 800 were used as the sample. An estimate of the median is $\hat\theta=`r med`$ and its standard error $SE(\hat\theta)=`r se`$.