---
title: "STAT840 Final Exam"
author: "Branden Lee"
date: "4/26/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 6

#### a)
```{r}
data <- c(186,181,176,149,184,190,158,139,175,148,152,11,41,153,190,157,131,149,
          135,132)

t0 <- (mean(data) - 165) / (sd(data) / sqrt(length(data)))
p_val <- pt(abs(t0), df=length(data) - 1, lower.tail=FALSE)
```
Note for the normal distribution the median $\theta$ is equal to the mean $\mu$ being a symmetric distribution. Then under $H_0$,
$$t_0=\frac{\overline X - \mu}{s/\sqrt{n}}\sim t_{n-1}$$
$p=P(t\ge |t_o|)=`r p_val`$. Hence at the 95% significance level we reject the null hypothesis, that is there is significant evidence that $\theta\ne 165$.

#### b)
```{r}
library(binaryLogic)

permutation_test <- function() {
  new_data <- data - 165
  T_obs <- mean(new_data)
  count <- 0
  for (b in 0:(2^20-1)) {
    bin <- as.binary(b)
    pi_b <- sapply(c(rep(0, 20-length(bin)), bin), function(x) 2*x-1)
    T_b <- mean(abs(new_data) * pi_b)

    if (abs(T_b) > abs(T_obs)) {
      count <- count + 1
    }

  }

  count / 2 ^ 20
}
permutation_test()
```
The permutation test results in a p value that is not significant at the 95% level, so the results of this test are different from that of part a.